Semantic Preserving Bijective Mappings for Expressions involving Special Functions between Computer Algebra Systems and Document Preparation Systems

### 2018-11-29 main.tex
made updates to main.tex

\noindent\textbf{Approach:} Bruce Miller at \gls*{nist} developed a collection of special \LaTeX{} macros that create links from mathematical symbols to their definitions in the \gls*{nist} \gls*{dlmf}. We are using these macros to perform rule-based translations between the formulae in the \gls*{dlmf} and \gls*{cas}. Moreover, we develop software to ease the creation of new rules and to discover inconsistencies.

\noindent\textbf{Findings:} We created 396 mappings and translated 58.8\% of the \gls*{dlmf} (2,405 expressions) successfully between \Maple{} and \gls*{dlmf}. For a significant percentage, the definitions in \Maple{} and the \gls*{dlmf} were different. Consequently, an atomic symbol in one system maps to a composite expression in the other system. The translator was also successfully used for automatic verification of mathematical online compendia and \gls*{cas}. Our evaluation techniques discovered two errors in the \gls*{dlmf} and one defect in \Maple.

@@

\section{Introduction}
A typical workflow of a scientist who writes a scientific publication is to use a \gls*{dps} to write the paper and one or more \glspl*{cas} for verification, analysis and visualization. Especially in the \gls*{stem} literature, \LaTeX{}
%\footnote{Note that technically \LaTeX{} is not a \gls*{dps} (\url{https://www.latex-project.org/about/}, seen 07/2017) like \textit{Microsoft Word}. However, since \LaTeX{} (an extension of \TeX) is the de facto standard for writing articles in \gls*{stem}, and we only focus on mathematical writing in this paper, we have categorized \LaTeX{} as a \gls*{dps} as well.} 
has become the de facto standard for writing scientific publications over the past 30 years~\parencites{Knuth}[559]{DigitalTypo}{LATEX:Standard}. \LaTeX{} enables printing of mathematical formulae in a structure similar to handwritten style. For example, consider the specific Jacobi polynomial~\parencite[Table 18.3.1]{NIST:DLMF}
\begin{equation}\label{eq:P}
P_n^{(\alpha , \beta)}(\cos(a\Theta)),
\end{equation}
where $n$ is a nonnegative integer, $\alpha, \beta > -1$, and $a,\Theta\in{\mathbf R}$.
This mathematical expression can be written in \LaTeX{} as
\begin{equation*}
\verb|P_n^{(\alpha,\beta)}(\cos(a\Theta))|.
\end{equation*}

@@

Translations from generic \LaTeX{} to \gls*{cas} are difficult to realize since the full semantic information is not easily constructed from the input. Bruce Miller at the \gls*{nist} has created a set of semantic \LaTeX{} macros~\parencite{DLMF:Macros}. Each macro ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf}. The \gls*{drmf} is an outgrowth of the \gls*{dlmf} with the goal to facilitate interaction among a community of mathematicians and scientists~\parencites{DRMF:14}{DRMF:15}. The \gls*{drmf} extends the set of semantic macros. These macros embed necessary semantic information into \LaTeX{} expressions. The macros may also contain $@$ symbols preceding the variables of the function. The number of the symbols is used to switch between different notation styles, e.g., $\cos(x)$ and $\cos x$. One example of such a macro is given in Table~\ref{tab:JacobiP-usecase} for the semantic \LaTeX{} representation for the Jacobi polynomial. The macros provide isolated access to important parts of the mathematical function, such as the arguments.

@@

\begin{figure}[ht]
    \centering
    \subfloat[The real part of arccotangent with a branch cut at ${[-\iunit\infty, -\iunit]}$, ${[\iunit,\iunit\infty]}$.]{
    		\includegraphics[width=0.45\textwidth]{acotCut1.jpg}
        \label{fig:acot-cut1}
    }
    \hspace{0.5cm}
    \subfloat[The real part of arccotangent with a branch cut at ${[-\iunit, \iunit]}$.]{
    		\includegraphics[width=0.45\textwidth]{acotCut2.jpg}
        \label{fig:acot-cut2}
    }
    \caption{Two plots for the real part of the arccotangent function with a branch cut at $[-\iunit\infty, -\iunit]$, $[\iunit,\iunit\infty]$ in Figure~\protect\subref{fig:acot-cut1} and at $[-\iunit, \iunit]$ in Figure~\protect\subref{fig:acot-cut2}, respectively. (Plotted with \Maple{} 2016)}
    \label{fig:acot-cut-compare}
\end{figure}

@@

Instead of using the term \textit{equivalent}, 
we introduce the terms \textit{appropriate} and \textit{inappropriate} translations. 
%{\bf [HSC: I don't understand how {\it equivalent} translations has anything at 
%all to do with {\it appropriate} or {\it inapppropriate} translations. Please clarify.]}.
We consider a translation as {\it appropriate}, when a numerical evaluation returns the 
same values in both concepts up to a numerical precision $|\epsilon|<<1$, for all possible 
points in the domain of the functions.
A translation is considered as {\it inappropriate}, when it is not {\it appropriate}.

For example, a translation such as
\begin{equation}\label{eq:cos-def}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|cos(z)|
\end{equation}
is \textit{appropriate}, while a translation such as
\begin{equation}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|sin(z)|
\end{equation}
is \textit{inappropriate}. Note that it is not always as easy as in this example to decide if a translation is {\it appropriate} or not. Therefore, this article also presents several validation techniques to automatically verify if a translation is {\it appropriate} or {\it inappropriate}. In addition to this terminology, we introduce \textit{direct translations}. Most mathematical objects in one system have a direct counterpart in other systems. However, later in the paper, we will explain that a translation from one specific mathematical object to its counterpart in the other system is not always \textit{appropriate}. Also, not every mathematical object has a counterpart in other systems. We call a translation to its counterpart \textit{direct}. For example, the translation~(\ref{eq:cos-def}) is \textit{direct}, while a translation to the definition of the cosine function
\begin{equation*}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|(exp(I*z)+exp(-I*z))/2|
\end{equation*}
is not a \textit{direct} translation even though it is {\it appropriate}.
Note that partial results of this paper have been published in~\parencite{CICM:Paper}.

\section{Related Work}\label{sec:related-work}
Since \LaTeX{} became the de facto standard for writing papers in mathematics, most of the \gls*{cas} provide simple functions to import and export mathematical \LaTeX{} expressions\footnote{The selected \gls*{cas} Maple, Mathematica, Matlab, and SageMath provide import and/or export functions for \LaTeX:\\
Maple, \url{http://www.maplesoft.com/support/help/Maple/view.aspx?path=latex} seen 06/2017;
Mathematica, \url{https://reference.wolfram.com/language/tutorial/GeneratingAndImportingTeX.html} seen 06/2017;
Matlab, \url{https://www.mathworks.com/help/symbolic/latex.html} seen 06/2017;
SageMath, \url{http://doc.sagemath.org/html/en/tutorial/latex.html} seen 06/2017.}. Those tools have two essential problems. They are only able to import simple mathematical expressions, where the semantics are unique. For example, the internal \LaTeX{} macro \verb|\frac| always indicates a fraction. However, for more complex expressions, e.g., the Jacobi polynomial in \Cref{tab:JacobiP-usecase}, the import functions fail. The second problem appears in the export tools. Mathematical expressions in \gls*{cas} are fully semantic. Otherwise the \gls*{cas} wouldn't be able to compute or evaluate the expressions. During the export process, the semantic information gets lost, because generic \LaTeX{} is not able to carry sufficient semantic information. Because of these two problems, an exported expression cannot be imported to the same system again in most cases (except for simple expressions such as those described above). Our tool should solve these problems and provide round-trip translations between \LaTeX{} and \gls*{cas}.

The semantics must be well known before an expression can be translated. There are two main approaches to solve that problem: (1) someone could specify the semantic information during the writing process (pre-defined semantics); and (2) the translator can find out the right semantic information in general mathematical expressions before it translates the expression. So-called \textit{interactive documents}\footnote{There is no adequate definition for what interactive documents are. However, this name is widely used to describe electronic document formats that allow for interactivity to change the content in real time.}, such as the \gls*{cdf}\footnote{Wolfram Research; \textit{Computable Document Format} (CDF); \url{http://www.wolfram.com/cdf/}, July 2011} by Wolfram Research, or \textit{worksheets} by \Maple{}, try to solve this problem with the approach (2) and allow one to embed semantic information into the input. Those complex document formats require specialized tools to show and work with the documents (Wolfram CDF Player, or \Maple{} for the \textit{worksheets}). The \JOBAD{} architecture~\parencite{JOBAD:orig} is able to create web-based interactive documents and uses \gls*{omdoc}~\parencite{OMDoc} to carry semantics. The documents can be viewed and edited in the browser. Those \JOBAD{-documents} also allow one to perform computations via \gls*{cas}. This gives one the opportunity to calculate, compute and change mathematical expressions directly in the document. The translation performs in the background, invisible to the user. Similar to the \JOBAD{} architecture, other interactive web documents exist, such as \textit{MathDox}~\parencite{MathDox} and \textit{The Planetary System}~\parencite{Planetary}.

Another approach tries to avoid translation problems by allowing computations directly via the \LaTeX{} compiler, e.g., \textit{LaTeXCalc}~\parencite{LatexCalc}. Those packages are limited to the abilities of the compiler and therefore are not as powerful as \gls*{cas}. A workaround for this case is \textit{sagetex}~\parencite{Sagetex}, which is a \LaTeX{} package interface for the open source \gls*{cas} \textit{sage}\footnote{An abbreviation for \textit{SageMath}.}. This package allows \textit{sage} commands in \TeX{-}files and uses \textit{sage} in the background to compute the commands. In this scenario, a writer still needs to manually translate expressions to the syntax of \textit{sage}, but it is possible to integrate \gls*{cas} expression directly into \TeX{} documents.

There exist two approaches for marking up mathematical \TeX/\LaTeX{} documents semantically with \TeX{} macros. Namely, \sTeX{}~\parencite{sTeX} developed by Kohlhase and the \Macro s developed by Miller~\parencite{DLMF:Macros}. This paper shows that it is possible to develop a context-free translation tool using the semantic macros introduced by these two projects. The goal of \sTeX{} was to markup the functional structure of mathematical documents so that they can be exported to the \gls*{omdoc} format. The macro set developed by Miller introduces new macros for special functions, orthogonal polynomials, and mathematical constants. Each of these macros ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf} or \gls*{drmf}. Therefore, we call these semantic macros \Macro s. These semantic macros are internally used in the \gls*{dlmf} and the \gls*{drmf}. We gave the \Macro{} set the favor for developing the translation engine because it provides \gls*{dlmf} definitions for a comprehensive number of functions. In contrast, \sTeX{} does not focus on the semantics of functions, is too complex to use, and defines diverse macros for symbols and concepts that \gls*{cas} usually does not support.

Miller also developed LaTeXML, a tool for converting \LaTeX{} expressions to MathML~\parencite{LaTeXML}. LaTeXML was used to generate the \gls*{dlmf} and is able to parse the \Macro s to content MathML. Even though many \gls*{cas} are able to import and export MathML, they fail for special functions. Schubotz et al.~recently performs benchmarks on several \LaTeX{} to MathML conversion tools, including LaTeXML, in~\parencite{DBLP:conf/jcdl/SchubotzGSMCG18}.

@@

\subsection{Different Sets of Defined Functions}
If a function is defined in one system but not in the other, we can translate the definition of the mathematical function. For example, the \textit{Gudermannian}~\parencite[(4.23.10)]{NIST:DLMF} $\Gudermannian{x}$ function is defined by
\begin{equation}\label{eq:gudermannian}
\Gudermannian{x} := \atan{\sinh{x}}, \quad x \in \Real,
\end{equation}
and linked to the semantic macro \verb|\Gudermannian| in the \gls*{dlmf} but does not exist in \Maple. We can perform a translation for the definition~\eqref{eq:gudermannian} instead of macro itself
\begin{equation}\label{eq:guder-trans}
\verb|\Gudermannian{x}| \overset{\langMaple}{\mapsto} \texttt{arctan(sinh(x))}.
\end{equation}

\vspace{-0.2cm}
Since translations such as these are nonintuitive, describing explanations become necessary for the translation process. A special logging function takes care of each translation and provides details after a successful translation process. \Cref{sec:forward-translation} explains this task further.

@@

\subsection{Positions of Branch Cuts}
In case of differences between defined branch cuts we can also use alternative translations to solve the problems. Consider the mentioned case of the arccotangent function~\parencite{Branches:acot} that has different positioned branch cuts in \Maple{} compared to the \gls*{dlmf} or \Mathematica{} definitions. As suggested by~\parencite{Branches:acot}, we can translate an alternative definition of the arccotangent function to avoid the branch cut issues. Considering the equations (23) and (25) in~\parencite{Branches:acot}, we can define three translations
\begin{eqnarray}
\verb|\acot@{z}| & \overset{\langMaple}{\mapsto} & \verb|arccot(z)|,\label{eq:acot-alternatives}\\
& \overset{\langMaple}{\mapsto} & \verb|arctan(1/z)|,\label{eq:acot-alternatives-1}\\
& \overset{\langMaple}{\mapsto} & \verb|I/2*ln((z-I)/(z+I))|.\label{eq:acot-alternatives-2}
\end{eqnarray}

\vspace{-0.2cm}
The position of the branch cut of the arccotangent function differs after the {\it direct} translation~\eqref{eq:acot-alternatives}, which may lead to incorrect calculations later on. The alternative translations~\eqref{eq:acot-alternatives-1} and~\eqref{eq:acot-alternatives-2} use other functions instead of the arccotangent function. The arctangent function~\eqref{eq:acot-alternatives-1} and the natural logarithm~(\ref{eq:acot-alternatives-2}) have the same positioned branch cuts as in the \gls*{dlmf} and in \Maple. Hence, translation~(\ref{eq:acot-alternatives-1}) solves the issue, as long as the user does not evaulate the function at $z = 0$, while translation~(\ref{eq:acot-alternatives-2}) solves the issue except at $z = -\iunit$. Note that none of the translations~(\ref{eq:acot-alternatives}-\ref{eq:acot-alternatives-2}) are {\it appropriate}.

\subsection{Insufficient Semantic Information}
Other problematic cases for translations are the \Macro s themselves. In some cases, they do not provide sufficient semantic information to perform translations. One example is the \textit{Wronskian} determinant. For two differentiable functions, the \textit{Wronskian} is defined as~\parencite[(1.13.4)]{NIST:DLMF}
\begin{equation*}
\mathscr{W}\{ w_1(z), w_2(z) \} = w_1(z)w_2'(z) - w_2(z)w_1'(z).
\end{equation*}
In semantic \LaTeX{} it is currently implemented using
\begin{equation}
	\verb|\Wronskian@{w_1(z), w_2(z)}|.
\end{equation}
A translation becomes unfeasible because the macro does not explicitly define the variable of differentiation for the functions $w_1$ and $w_2$. For a correct translation, the \gls*{cas} needs to be aware of the used variable $z$. We solved this issue by creating a new macro, e.g.,
\begin{equation}
	\verb|\Wron{z}@{w_1(z)}{w_2(z)}|.
\end{equation}
This example demonstrates that the \Macro s are still a work in progress and are getting constantly updated.

\subsection{Ambiguous Expressions}
A similar problem is multiplications since they are rarely explicitly marked in \LaTeX{} expressions, e.g., scientists using whitespaces to indicate multiplications rather than using \verb|\cdot| or similar symbols. For such problems, we introduced a new macro \verb|\idot| for an invisible multiplication symbol (this macro will not be rendered). Since this macro is newly introduced by contributers of the \gls*{drmf} team, and automatic conversion of existing equations is difficult, none of the equations in the \gls*{dlmf} use this macro. As a consequence, the translator has some simple rules to perform translations without explicitly marked translations with \verb|\idot|.

@@

Another more questionable translation decision is alphanumerical expressions. As explained in~\Cref{tab:allTypesTable}, the \gls*{pom}-tagger handles strings of letters and numbers differently, depending on the order of the symbols. The reason is that an expression such as `$4b$' is usually considered to be a multiplication of $4$ and `$b$,' while `$b4$' looks like indexing `$b$' by $4$. While the first example produces two nodes, namely $4$ and `$b$', the second example `$b4$' produces just a single alphanumerical node in the \gls*{pom-pt}. The translator interprets alphanumerical expressions as multiplications for two reasons: (1) we would assume that the inputs `$4b$' and `$b4$' are mathematically equivalent; and (2) it is more common in mathematics to use single letter names for variables~\parencite{Notation:History}. Therefore we make the following definitions

@@

In general, the translator is designed to find a work-around for disambiguating expressions. Only if there is no way to solve the ambiguity with the defined rules, the translation process stops.

\section{The Translator}
The translator analyzes a parse tree to perform translations. For generating a parse tree of \LaTeX{} expressions, the translator uses the \gls*{pom}-Tagger~\parencite{POM-Tagger}\footnote{Named according to the Part-of-Speech-Taggers in \gls*{nlp}.}. \gls*{cas} define their own syntax parser. We were able to use \Maple's internal data structure to get a parse tree of the input. \Cref{sec:forward-translation} and \Cref{sec:backward-translation} will explain the parsing and translation process in detail.

All translations are defined by a library (\gls*{csv} and \gls*{json} files) that defines translation patterns for each function and symbol. The pattern uses \verb|$i| as a placeholder to define the positions of the arguments. For example, the translation patterns for the Jacobi polynomial are illustrated in~\Cref{tab:placeholder_ex2}.

@@

There are unevaluation quotes implemented to avoid evaluations on input expression. Table~\ref{tab:unevaluation-quotes} gives an example how those unevaluation quotes work.

>>>>

\noindent\textbf{Approach:} Bruce Miller at the \gls*{nist} developed a collection of special \LaTeX{} macros that create links from mathematical symbols to their definitions in the \gls*{nist} \gls*{dlmf}. We are using these macros to perform rule-based translations between the formulae in the \gls*{dlmf} and \gls*{cas}. Moreover, we develop software to ease the creation of new rules and to discover inconsistencies.

\noindent\textbf{Findings:} We created 396 mappings and translated 58.8\% of \gls*{dlmf} formulae (2,405 expressions) successfully between \Maple{} and \gls*{dlmf}. For a significant percentage, the special function definitions in \Maple{} and the \gls*{dlmf} were different. Consequently, an atomic symbol in one system maps to a composite expression in the other system. The translator was also successfully used for automatic verification of mathematical online compendia and \gls*{cas}. Our evaluation techniques discovered two errors in the \gls*{dlmf} and one defect in \Maple.

@@

\section{Introduction}
A typical workflow of a scientist who writes a scientific publication is to use a \gls*{dps} to write the paper and one or more \glspl*{cas} for verification, analysis and visualization. Especially in the \gls*{stem} literature, \LaTeX{}
%\footnote{Note that technically \LaTeX{} is not a \gls*{dps} (\url{https://www.latex-project.org/about/}, seen 07/2017) like \textit{Microsoft Word}. However, since \LaTeX{} (an extension of \TeX) is the de facto standard for writing articles in \gls*{stem}, and we only focus on mathematical writing in this paper, we have categorized \LaTeX{} as a \gls*{dps} as well.} 
has become the de facto standard for writing scientific publications over the past 30 years~\parencites{Knuth}[559]{DigitalTypo}{LATEX:Standard}. \LaTeX{} enables printing of mathematical formulae in a structure similar to handwritten style. For example, consider the specific Jacobi polynomial~\parencite[Table 18.3.1]{NIST:DLMF}
\begin{equation}\label{eq:P}
P_n^{(\alpha , \beta)}(\cos(a\Theta)),
\end{equation}
where $n$ is a nonnegative integer, $\alpha, \beta > -1$, and $a,\Theta\in \Real$.
This mathematical expression can be written in \LaTeX{} as
\begin{equation*}
\verb|P_n^{(\alpha,\beta)}(\cos(a\Theta))|.
\end{equation*}

@@

Translations from generic \LaTeX{} to \gls*{cas} are difficult to realize since the full semantic information is not easily constructed from the input. Bruce Miller at the \gls*{nist} has created a set of semantic \LaTeX{} macros~\parencite{DLMF:Macros}. Each macro ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf}. The \gls*{drmf} is an outgrowth of the \gls*{dlmf} with the goal to facilitate interaction among a community of mathematicians and scientists~\parencites{DRMF:14}{DRMF:15}. The \gls*{drmf} extends the set of semantic macros. These macros embed necessary semantic information into \LaTeX{} expressions. The macros may also contain $@$ symbols preceding the variables of the function. The number of $@$ symbols is used to switch between different notation styles, e.g., $\cos(x)$ and $\cos x$. One example of such a macro is given in Table~\ref{tab:JacobiP-usecase} for the semantic \LaTeX{} representation for the Jacobi polynomial. The macros provide isolated access to important parts of the mathematical function, such as the arguments.

@@

\begin{figure}[ht]
    \centering
    \subfloat[The real part of arccotangent with a branch cut at ${[-\iunit\infty, -\iunit]}$, ${[\iunit,\iunit\infty]}$.]{
    		\includegraphics[width=0.45\textwidth]{acotCut1.jpg}
        \label{fig:acot-cut1}
    }
    \hspace{0.5cm}
    \subfloat[The real part of arccotangent with a branch cut at ${[-\iunit, \iunit]}$.]{
    		\includegraphics[width=0.45\textwidth]{acotCut2.jpg}
        \label{fig:acot-cut2}
    }
    \caption{Two plots of the real part for the arccotangent function with a branch cut at $[-\iunit\infty, -\iunit]$, $[\iunit,\iunit\infty]$ in Figure~\protect\subref{fig:acot-cut1} and at $[-\iunit, \iunit]$ in Figure~\protect\subref{fig:acot-cut2}, respectively. (Plotted with \Maple{} 2016)}
    \label{fig:acot-cut-compare}
\end{figure}

@@

In order to clarify the notion of \textit{equivalence} (or lack thereof) in our context of translations,
we introduce the terms \textit{appropriate} and \textit{inappropriate} translations. 
%{\bf [HSC: I don't understand how {\it equivalent} translations has anything at 
%all to do with {\it appropriate} or {\it inapppropriate} translations. Please clarify.]}.
We consider a translation to be {\it appropriate}, when a numerical evaluation returns the 
same values in both concepts up to a numerical precision $|\epsilon|\ll 1$, for all possible 
points in specified domains for the functions.
A translation is considered as {\it inappropriate}, when it is not {\it appropriate}.

For example, a translation such as
\begin{equation}\label{eq:cos-def}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|cos(z)|
\end{equation}
is \textit{appropriate}, while a translation such as
\begin{equation}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|sin(z)|
\end{equation}
is \textit{inappropriate}. Note that it is not always as easy as in this example to decide if a translation is {\it appropriate} or not. Therefore, this article also presents several validation techniques to automatically verify if a translation is {\it appropriate} or {\it inappropriate}.

In addition, we also introduce the notion of \textit{direct} translations. Most mathematical objects in one system have a direct counterpart in other systems. However, later in the paper, we will explain that a translation from one specific mathematical object to its counterpart in the other system is not always \textit{appropriate}. Also, not every mathematical object has a counterpart in other systems. We call a translation to its counterpart \textit{direct}. For example, the translation~(\ref{eq:cos-def}) is \textit{direct}, while a translation to the definition of the cosine function
\begin{equation*}
\verb|\cos@{z}| \overset{\langMaple}{\mapsto} \verb|(exp(I*z)+exp(-I*z))/2|
\end{equation*}
is not a \textit{direct} translation even though it is {\it appropriate}.
Note that partial results of this paper have been published in~\parencite{CICM:Paper}.

\section{Related Work}\label{sec:related-work}
Since \LaTeX{} became the de facto standard for writing papers in mathematics, most \gls*{cas} provide simple functions to import and export mathematical \LaTeX{} expressions\footnote{The selected \gls*{cas} Maple, Mathematica, Matlab, and SageMath provide import and/or export functions for \LaTeX:\\
Maple, \url{http://www.maplesoft.com/support/help/Maple/view.aspx?path=latex} seen 06/2017;
Mathematica, \url{https://reference.wolfram.com/language/tutorial/GeneratingAndImportingTeX.html} seen 06/2017;
Matlab, \url{https://www.mathworks.com/help/symbolic/latex.html} seen 06/2017;
SageMath, \url{http://doc.sagemath.org/html/en/tutorial/latex.html} seen 06/2017.}. Those tools have two essential problems. They are only able to import simple mathematical expressions, where the semantics are unique. For example, the internal \LaTeX{} macro \verb|\frac| always indicates a fraction. However, for more complex expressions, e.g., the Jacobi polynomial in \Cref{tab:JacobiP-usecase}, the import functions fail. The second problem appears in the export tools. Mathematical expressions in \gls*{cas} are fully semantic. Otherwise the \gls*{cas} wouldn't be able to compute or evaluate the expressions. During the export process, the semantic information gets lost, because generic \LaTeX{} is not able to carry sufficient semantic information. Because of these problems, an exported expression cannot be imported to the same system again in most cases (except for simple expressions such as those described above). Our tool attempts to solve these problems and provide round-trip translations between \LaTeX{} and \gls*{cas}.

The semantics must be well-known before an expression can be translated. There are two main approaches to solve that problem:~(1) someone could specify the semantic information during the writing process (pre-defined semantics); and (2) the translator can determine the correct semantic information in general mathematical expressions before it translates the expression. So-called \textit{interactive documents}\footnote{There is no adequate definition for what interactive documents are. However, this name is widely used to describe electronic document formats that allow for interactivity to change the content in real time.}, such as the \gls*{cdf}\footnote{Wolfram Research; \textit{Computable Document Format} (CDF); \url{http://www.wolfram.com/cdf/}, July 2011} by Wolfram Research, or \textit{worksheets} by \Maple{}, try to solve this problem with the approach (2) and allow one to embed semantic information into the input. Those complex document formats require specialized tools to show and work with the documents (Wolfram CDF Player, or \Maple{} for the \textit{worksheets}). The \JOBAD{} architecture~\parencite{JOBAD:orig} is able to create web-based interactive documents and uses \gls*{omdoc}~\parencite{OMDoc} to carry semantics. The documents can be viewed and edited in the browser. Those \JOBAD{-documents} also allow one to perform computations via \gls*{cas}. This gives one the opportunity to calculate, compute and change mathematical expressions directly in the document. The translation performs in the background, invisible to the user. Similar to the \JOBAD{} architecture, other interactive web documents exist, such as \textit{MathDox}~\parencite{MathDox} and \textit{The Planetary System}~\parencite{Planetary}.

Another approach tries to avoid translation problems by allowing computations directly via the \LaTeX{} compiler, e.g., \textit{LaTeXCalc}~\parencite{LatexCalc}. Those packages are limited to the abilities of the compiler and therefore are not as powerful as \gls*{cas}. A workaround for this case is \textit{sagetex}~\parencite{Sagetex}, which is a \LaTeX{} package interface for the open source \gls*{cas} \textit{sage}\footnote{An abbreviation for \textit{SageMath}.}. This package allows \textit{sage} commands in \TeX{-}files and uses \textit{sage} in the background to compute the commands. In this scenario, a writer still needs to manually translate expressions to the syntax of \textit{sage}, but it is possible to integrate \gls*{cas} expressions directly into \TeX{} documents.

There exist two approaches for marking up mathematical \TeX/\LaTeX{} documents semantically with \TeX{} macros. Namely, \sTeX{}~\parencite{sTeX} developed by Kohlhase and the \Macro s developed by Miller~\parencite{DLMF:Macros}. This paper shows that it is possible to develop a context-free translation tool using the semantic macros introduced by these two projects {\bf [HSC: which two projects? Please clarify.]}. The goal of \sTeX{} is to markup the functional structure of mathematical documents so that they can be exported to the \gls*{omdoc} format. The macro functionality developed by Miller introduces new macros for special functions, orthogonal polynomials, and mathematical constants. Each of these macros ties specific character sequences to a well-defined mathematical object and is linked with the corresponding definition in the \gls*{dlmf} or \gls*{drmf}. Therefore, we call these semantic macros \Macro s. These semantic macros are internally used in the \gls*{dlmf} and the \gls*{drmf}. We gave the \Macro{} set the favor for developing the translation engine because it provides \gls*{dlmf} definitions for a comprehensive number of functions. In contrast, \sTeX{} does not focus on the semantics of functions, is often complex to use, and defines diverse macros for symbols and concepts that \gls*{cas} usually does not usually support.

Miller also developed LaTeXML, a tool for converting \LaTeX{} expressions to MathML~\parencite{LaTeXML}. LaTeXML is used to generate the \gls*{dlmf} and is able to parse the \Macro s to generate content MathML. Even though many \gls*{cas} are able to import and export MathML, they fail for special functions. Schubotz and collaborators recently performed benchmarks on several \LaTeX{} to MathML conversion tools, including LaTeXML, in~\parencite{DBLP:conf/jcdl/SchubotzGSMCG18}.

@@

\subsection{Different Sets of Defined Functions}
If a function is defined in one system but not in the other, sometimes we can easily translate the definition of the mathematical function. For example, the \textit{Gudermannian}~\parencite[(4.23.10)]{NIST:DLMF} $\Gudermannian{x}$ function is defined by
\begin{equation}\label{eq:gudermannian}
\Gudermannian{x} := \atan{\sinh{x}}, \quad x \in \Real,
\end{equation}
and linked to the semantic macro \verb|\Gudermannian| in the \gls*{dlmf} but does not exist in \Maple. We can perform a translation for the definition~\eqref{eq:gudermannian} instead of macro itself
\begin{equation}\label{eq:guder-trans}
\verb|\Gudermannian{x}| \overset{\langMaple}{\mapsto} \texttt{arctan(sinh(x))}.
\end{equation}

\vspace{-0.2cm}
Since translations such as these are nonintuitive, describing explanations become necessary for the translation process. A special logging function takes care {\bf [HSC: ``takes care'' is too informal. Use a different phrase or word.]} of each translation and provides details after a successful translation process. \Cref{sec:forward-translation} explains this task further.

@@

\subsection{Positions of Branch Cuts}
In case of differences between defined branch cuts, we can also use alternative translations to solve the problems. Consider the mentioned case of the arccotangent function~\parencite{Branches:acot} that has different positioned branch cuts in \Maple{} as compared to the \gls*{dlmf} or \Mathematica{} definitions. As suggested by~\parencite{Branches:acot}, we can translate an alternative definition of the arccotangent function to avoid the branch cut issues. Considering 
\parencite[(23), (25)]{Branches:acot}, we can define three translations
\begin{eqnarray}
\verb|\acot@{z}| & \overset{\langMaple}{\mapsto} & \verb|arccot(z)|,\label{eq:acot-alternatives}\\
& \overset{\langMaple}{\mapsto} & \verb|arctan(1/z)|,\label{eq:acot-alternatives-1}\\
& \overset{\langMaple}{\mapsto} & \verb|I/2*ln((z-I)/(z+I))|.\label{eq:acot-alternatives-2}
\end{eqnarray}

\vspace{-0.2cm}
The position of the branch cut of the arccotangent function differs after the {\it direct} translation~\eqref{eq:acot-alternatives}, which may lead to incorrect calculations later on. The alternative translations~\eqref{eq:acot-alternatives-1} and~\eqref{eq:acot-alternatives-2} use other functions instead of the arccotangent function. The arctangent function~\eqref{eq:acot-alternatives-1} and the natural logarithm~(\ref{eq:acot-alternatives-2}) have the same positioned branch cuts as in the \gls*{dlmf} and in \Maple. Hence, translation~(\ref{eq:acot-alternatives-1}) solves this issue as long as the user does not evaulate the function at $z = 0$, while translation~(\ref{eq:acot-alternatives-2}) solves the issue except at $z=-\iunit$. Note that none of the translations~(\ref{eq:acot-alternatives}-\ref{eq:acot-alternatives-2}) are {\it appropriate}.

\subsection{Insufficient Semantic Information}
Other problematic cases for translations are the \Macro s themselves. In some cases, they do not provide sufficient semantic information to perform translations. One example is the \textit{Wronskian} determinant. For two differentiable functions $w_1$, $w_2$, the \textit{Wronskian} is defined as~\parencite[(1.13.4)]{NIST:DLMF}
\begin{equation*}
\mathscr{W}\{ w_1(z), w_2(z) \} = w_1(z)w_2'(z) - w_2(z)w_1'(z).
\end{equation*}
In semantic \LaTeX{}, it is currently implemented using
\begin{equation}
	\verb|\Wronskian@{w_1(z), w_2(z)}|.
\end{equation}
This translation is unfeasible because the macro does not explicitly define the variable of differentiation for the functions $w_1$, $w_2$. For a correct translation, the \gls*{cas} needs to be aware of the variable of differentiation $z$. We solved this issue by creating a new macro \verb|\Wron|, e.g.,
\begin{equation}
	\verb|\Wron{z}@{w_1(z)}{w_2(z)}|.
\end{equation}
This example demonstrates that the \Macro s are still a work in progress and further updates are sometimes
necessary in order to further encapsulate critical semantic information {\bf [HSC: I changed this sentence. Check to make sure that it is still correct.]}.

\subsection{Potentially Ambiguous Expressions}

{\bf [HSC: This section needs to start out with a short overview paragraph describing the issue instead of an example. Also, I changed the title of the section.]}

A similar problem is multiplications since they are rarely explicitly marked in \LaTeX{} expressions, e.g., scientists using whitespace to indicate multiplications rather than using \verb|\cdot| or similar symbols. For such problems, we introduced a new macro \verb|\idot| for an invisible multiplication symbol (this macro will not be rendered). Since this macro is newly introduced by contributers of the \gls*{drmf} team, and automatic conversion of existing equations is difficult, none of the equations in the \gls*{dlmf} use this macro. As a consequence, the translator has some simple 
rules for performing translations without explicitly marking multiplication translations with \verb|\idot|.

@@

Another more questionable translation decision is alphanumerical expressions. As explained in~\Cref{tab:allTypesTable}, the \gls*{pom}-tagger handles strings of letters and numbers differently depending on the order of the symbols. The reason is that an expression such as `$4b$' is usually considered to be a multiplication of $4$ and `$b$,' while `$b4$' 
displays like indexing `$b$' by $4$. While the first example produces two nodes, namely $4$ and `$b$', the second example `$b4$' produces just a single alphanumerical node in the \gls*{pom-pt}. The translator interprets alphanumerical expressions as multiplications for two reasons:~(1) we would assume that the inputs `$4b$' and `$b4$' are mathematically equivalent; and (2) it is more common in mathematics to use single letter names for variables~\parencite{Notation:History}. Therefore we have used rules as follows {\bf [HSC: I changed the wording here, definition$\to$rules. Is this ok? Otherwise further clarify.]}

@@

In general, the translator is designed to find a work-around for disambiguating expressions. 
If there is no way to solve a potential ambiguity with defined rules, then we stop the translation process
{\bf [HSC: I changed the wording here also. Is this ok?]}.

\section{The Translator}
The translator analyzes a parse tree to perform translations. For generating a parse tree of \LaTeX{} expressions, the translator uses the \gls*{pom}-Tagger~\parencite{POM-Tagger}\footnote{Named according to the Part-of-Speech-Taggers in \gls*{nlp}.}. \gls*{cas} define their own syntax parser. We were able to use \Maple's internal data structure to 
obtain {\bf [HSC: I changed get$\to$obtain.]} a parse tree of the input. \Cref{sec:forward-translation} and \Cref{sec:backward-translation} will explain the parsing and translation process in detail.

All translations are defined by a library (\gls*{csv} and \gls*{json} files) that define translation patterns for each function and symbol. The pattern uses \verb|$i| as a placeholder used to describe {\bf [HSC: I used describe instead of define. Is this ok? Otherwise use a different more descriptive term.]} the positions of the arguments. For example, the translation patterns for the Jacobi polynomial are illustrated in~\Cref{tab:placeholder_ex2}.

@@

There are unevaluation quotes implemented to avoid evaluations on input expressions. Table~\ref{tab:unevaluation-quotes} gives an example how unevaluation quotes work.
